---
title: "DADA2 pipeline — 16S amplicon (HiSeq, MiSeq)"
author: "Simon Yersin"
date: '2026-06-09'
output: pdf_document
---

IMPORTANT NOTE: This pipeline is for DNA sequenced on MiSeq or HiSeq Illumina sequencing systems! 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Starting point ---

Our starting point is a set of Illumina-sequenced paired-end fastq files that have been split (or “demultiplexed”) by sample and from which the barcodes/adapters have already been removed
The end product is an amplicon sequence variant (ASV) table, a higher-resolution analogue of the traditional OTU table, which records the number of times each exact amplicon sequence variant was observed in each sample
We also assign taxonomy to the output sequences

This workflow assumes that your sequencing data meets certain criteria:
DNA sequenced on MiSeq or HiSeq Illumina sequencing systems.
Samples have been demultiplexed, i.e. split into individual per-sample fastq files.
Non-biological nucleotides have been removed, e.g. primers, adapters, linkers, etc.
If paired-end sequencing data, the forward and reverse fastq files contain reads in matched order.

IF single library, remove the reverese read and do not have to merge

Tutorial:
https://benjjneb.github.io/dada2/tutorial.html

Note: 
If multiple batches are processed together with DADA2, the following steps must be performed separately per batch: upload files, filter and trim, learn error rate, dereplication, sample inference, merging and construct sequence table.
Once a sequence table is constructed for each batch, they can be merged and the following steps can be done in common: remove chimeras and assign taxonomy.
IMPORTANT: all batches have to have the same region amplified! 

# Loading libraries
```{r library}
library(dada2)
library(ggplot2)
library(dplyr)
```

# Working directory and path
```{r set_path}
#Set theme for ggplots
theme_set(theme_bw())

# Set your working directory
# First check which wd you are working in with:
getwd()
# Copy and paste path in the following to set your working directory:
setwd("/Users/admin/Documents/PATH_TO_PROJECT")

# Save path to fastq file folder after unzipping
path <- "/Users/admin/Documents/PATH_TO_FASTQ_FILES"
# Control files
list.files(path)
```

# Upload files
```{r load_files}
# Check for specific name in fastq for forward and reverse read
# Example fnFs: R1_001.fastq.gz fnRs: R2_001.fastq.gz
fnFs <- sort(list.files(path, pattern="_L001_R1_001.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_L001_R2_001.fastq.gz", full.names = TRUE))

# Extract sample names
# See files for pattern, ex from AFRIBIOTA fastq: _L001
sample.namesF<- sapply(strsplit(basename(fnFs), "_L001_R1_001"), `[`, 1)
sample.namesR <- sapply(strsplit(basename(fnRs), "_L001_R2_001"), `[`, 1)

#Verify if same number of forward and reverse:
length(fnFs) #146
length(fnRs) #146
intersect <- intersect(sample.namesF, sample.namesR)
sample.namesR [! sample.namesR %in% sample.namesF] # these are in the reverse but not in the forward --> NONE
sample.namesF [! sample.namesF %in% sample.namesR] # these are in the forward but not in the reverse --> NONE
```

# Quality plots
```{r quality_profiles}
# plot quality profiles of the forward and reverse reads 
plotQualityProfile(fnFs[1:4]) 
plotQualityProfile(fnRs[1:4])
# Allow to assess were the quality drops on the forward and reverse reads. These positions inform the settings of truncLen in filterAndTrim().
```

# Filter and trim
```{r filter_and_trim}
#Place filter files in filtered subdirectory
filtFs <- file.path(path, "dada2_filtered", paste0(sample.namesF, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "dada2_filtered", paste0(sample.namesR, "_R_filt.fastq.gz"))

names(filtFs) <- sample.namesF
names(filtRs) <- sample.namesR

# Check parameters
# the more aggressive you are with truncation, the more reads will be retained,
# but you still need to merge, so consider how much read overlap remains after truncation
# Might have to remove verbose and MatchIDs
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, 
                     truncLen=c(240,220), 
                     maxN=0, truncQ=c(2,2), maxEE=c(2,2),
                     rm.phix=TRUE, verbose=TRUE, matchIDs=TRUE,
                     compress=TRUE, multithread=TRUE) 
# Check retained reads
retained <- as.data.frame(out)
retained$percentage_retained <- retained$reads.out/retained$reads.in*100
View(retained)

# Or check quality profile again
plotQualityProfile(filtFs[1:4])
plotQualityProfile(filtRs[1:4])

# Samples with fewer than 1000 reads after filtering are dropped before error
# learning. This is safer than filtering after DADA2 inference, because very
# low-read samples can distort the error model.
samples_to_keep_filter <- retained$reads.out >= 1000
cat("Samples dropped at filter step:", sum(!samples_to_keep_filter), "\n")
```

# Error rate
```{r learn_error}
#Learn error rates
errF <- learnErrors(filtFs[samples_to_keep_filter], multithread=TRUE)
errR <- learnErrors(filtRs[samples_to_keep_filter], multithread=TRUE)

#Plot error
plotErrors(errF, nominalQ=TRUE)
plotErrors(errR, nominalQ=TRUE)
```

# Dereplication
```{r derep}
derepFs <- derepFastq(filtFs[samples_to_keep_filter], verbose=TRUE)
derepRs <- derepFastq(filtRs[samples_to_keep_filter], verbose=TRUE)

# Name the derep-class objects by the sample names
names(derepFs) <- sample.namesF[samples_to_keep_filter]
names(derepRs) <- sample.namesR[samples_to_keep_filter]
```

# Sample inference
```{r dada}
#DADA2 infers sample sequences exactly and resolves differences of as little as 1 nucleotide
dadaFs <- dada(derepFs, err=errF, multithread=TRUE)
dadaRs <- dada(derepRs, err=errR, multithread=TRUE)

#Inspecting the returned dada-class object:
dadaFs[[1]]
dadaRs[[1]]
```

#Remove low-sequence sample
```{r, sample_filter}
# After denoising, verify that all samples retained sufficient reads.
# This tracks unique sequences at each stage as an additional check.
getN <- function(x) sum(getUniques(x))

track_pre <- cbind(
  sapply(derepFs, getN),
  sapply(derepRs, getN),
  sapply(dadaFs,  getN),
  sapply(dadaRs,  getN))

colnames(track_pre) <- c("derepF", "derepR", "dadaF", "dadaR")
rownames(track_pre) <- sample.namesF[samples_to_keep_filter]

# We now use it only as a secondary check — primary filtering happened at filterAndTrim.
samples_to_keep <- track_pre[, 4] > 100
cat("Samples dropped at post-DADA2 check:", sum(!samples_to_keep), "\n")
track_pre[, , drop = FALSE]
```


# Merging
```{r merge}
# Merge forward and reverse denoised reads to obtain full-length amplicon sequences.
# Pairs must overlap by at least minOverlap bases with at most maxMismatch mismatches.

# --- VERIFY OVERLAP GEOMETRY BEFORE RUNNING ---
# Expected overlap = truncLenF + truncLenR - amplicon_length
# For V3 (~253 bp):  220 + 200 - 253 = 167 bp — merging should work well.
# If you change truncLen values, recalculate this before proceeding.
# If overlap is <20 bp, merging will fail for most reads.


# Merge forward and reverse reads to obtain the full denoised sequences
# Skip this step if using fastq with only forward reads
mergers <- mergePairs(
  dadaFs[samples_to_keep],
  derepFs[samples_to_keep],
  dadaRs[samples_to_keep],
  derepRs[samples_to_keep],
  minOverlap = 20,    # minimum overlap in bp; default is 12, 20 is safer
  maxMismatch = 0,    # no mismatches in the overlap region (default, keep this)
  verbose = TRUE
)


# The mergers object is a list of data.frames from each sample. 
# Each data.frame contains the merged $sequence, its $abundance, and the indices of the $forward and $reverse sequence variants that were merged. 

# Inspect the merger data frame from the first sample
# 'accept' column: TRUE means the pair was successfully merged
head(mergers[[1]])
```

# Construct sequence table
```{r seqtab}
seqtab <- makeSequenceTable(mergers)

#Check table dimension
dim(seqtab)

#Inspect districution of sequence lengths
table(nchar(getSequences(seqtab)))

# Inspect the distribution of merged sequence lengths
# For V3, expect a tight peak around 253 bp
# For V3-V4, expect a peak around 460 bp
seq_length_dist <- table(nchar(getSequences(seqtab)))
seq_length_dist

#View Sequence Length Distribution Post-Merging
#most useful with merged data. this plot will not show you much for forward reads only, which should have a uniform length distribution.
length.histogram <- as.data.frame(seq_length_dist) #tabulate sequence length distribution
plot(x=length.histogram[,1], y=length.histogram[,2],
     xlab = "Sequence length (bp)", ylab = "Number of ASVs",
     main = "Merged ASV length distribution") #view length distribution plot

## If different datasets (fastq files from different studies) are used, 
# the above pipeline need to be done for each dataset separatly
# Then the seqtab can be merged with: 
#seqtab <- mergeSequenceTables(tables = list(seqtab_1, seqtab_2, seqtab_3))

# then the rest of the pipeline can be done on the merged seqtab
```

```{r seqtab_length_filter}
# --- ADDED: amplicon length filter ---
# Retain only sequences within the expected length range for your amplicon.
# This removes off-target products (primer dimers, chimeric merges of wrong length)
# that survived to this point.
#
# For V3 (341F/806R region, ~253 bp after primer removal):
#   use seq(252, 255) as a generous window around the peak
#   Adjust based on your length distribution plot above.
#
# For V3-V4 (341F/806R, ~460 bp):
#   use seq(440, 480)

seqtab <- seqtab[, nchar(colnames(seqtab)) %in% seq(250, 256)]
cat("ASVs retained after length filter:", ncol(seqtab), "\n")
```


Comment: the removal low-read-count singleton ASVs step was deleted from this pipeline

# Remove chimeras
```{r}
# Here we remove "bimeras" or chimeras with two sources
# look at "method" to decide which type of pooling you'd like to use when judging each sequence as chimeric or non-chimeric
# Possible methods: "consensus" ; "pooled"
#this step can take a few minutes to a few hours, depending on the size of your datas
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE) 

dim(seqtab.nochim)
# Proportion of reads retained after chimera removal
# Should typically be >90% — you may lose many ASVs but should lose few reads
prop_retained <- sum(seqtab.nochim)/sum(seqtab)
cat(sprintf("Proportion of reads retained after chimera removal: %.1f%%\n",
            prop_retained * 100))
```

# Track reads
```{r track_reads}
# Look at the number of reads that made it through each step in the pipeline
getN <- function(x) sum(getUniques(x))

# IF ran the optional, run the following:
track <- cbind(out[samples_to_keep_filter, ][samples_to_keep, ],
               sapply(dadaFs[samples_to_keep], getN), 
               sapply(dadaRs[samples_to_keep], getN), 
               sapply(mergers, getN),
               rowSums(seqtab.nochim))


# If processing only a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)

# Adding percent_chimeras in track
track <- cbind(track, round(100-track[,6]/track[,5]*100,2))
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged",
                     "nochimeras","percent_chimeras")


rownames(track) <- sample.namesF[samples_to_keep_filter][samples_to_keep]

# Save output from sequence table construction steps
# Read retention table
write.table(track, "Tables/read_retention_table_PROJECT_NAME.txt", quote=F, row.names=T, col.names=T, sep="\t")
```

# Assign taxonomy
The SILVA v138.2 database is used in a **two-step approach**:

1. `assignTaxonomy()` — assigns Kingdom through Genus using a naïve Bayesian classifier.
2. `addSpecies()` — adds species-level assignments by exact-match to reference sequences.
```{r}
# Note: time consuming if you have a large dataset
  # Saving the sequences as a fasta file (with writeFasta) and using QIIME's taxonomy assignment command will save you time
  # Slightly less accurate than the dada2 package's taxonomy assignment function.
# Using the Silva database

# Silva database: https://benjjneb.github.io/dada2/training.html
# download the silva_nr99_v138.2_toGenus_trainset.fa.gz
# Assign to Genus
taxa <- assignTaxonomy(seqtab.nochim, "db/silva_nr99_v138.2_toGenus_trainset.fa.gz", multithread=TRUE, tryRC = TRUE)

# The dada2 package also implements a method to make species level assignments based on exact matching between ASVs and sequenced reference strains. Recent analysis suggests that exact matching (or 100% identity) is the only appropriate way to assign species to 16S gene fragments. Currently, species-assignment training fastas are available for the Silva and RDP 16S databases. To follow the optional species addition step, download the silva_v138.2_assignSpecies.fa.gz file, and place it in the directory with the fastq files. Using: 
taxa <- addSpecies(taxa, "db/silva_v138.2_assignSpecies.fa.gz")


# Label unassigned Kingdom-level ASVs
# These may be eukaryotic, archaeal, or true unknowns
NAs <- which(is.na(taxa[, 1]))
taxa[NAs, 1] <- "Unassigned"
cat("Unassigned at Kingdom level:", length(NAs), "\n")

# Quick summary of assignment rates by rank
cat("\nAssignment rate by rank:\n")
for (i in 1:ncol(taxa)) {
  assigned <- sum(!is.na(taxa[, i]))
  cat(sprintf("  %s: %d / %d (%.1f%%)\n",
              colnames(taxa)[i], assigned, nrow(taxa),
              assigned / nrow(taxa) * 100))
}
```

```{r export_tables}
# Create output directory if it does not exist
dir.create("Tables", showWarnings = FALSE)

# --- Step 1: add ASV labels and sequence column to the taxonomy table ---
# At this point taxa is a matrix (from assignTaxonomy / addSpecies).
# We convert to data.frame temporarily to use dplyr::mutate.
taxa <- taxa %>%
  as.data.frame() %>%
  mutate(ASV = paste0("ASV", seq(1, nrow(taxa))),  # short label: ASV1, ASV2, ...
         SEQ = rownames(taxa))                       # full sequence retained here

# Assign short ASV labels as row names
rownames(taxa) <- taxa$ASV

# Sanity check: confirm no duplicate ASV names were created
stopifnot(sum(duplicated(rownames(taxa))) == 0)
cat("No duplicate ASV names — OK\n")

# --- Step 2: build the correspondence table ---
# This maps ASV identifier <-> full sequence for future reference.
# Keep this file alongside your results.
correspondence <- taxa[, c("ASV", "SEQ")]
write.table(correspondence,
            "Tables/correspondence_table_PROJECT_NAME.txt",
            sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE)

# --- Step 3: convert taxa back to a matrix and drop the helper columns ---
# Columns 8 (ASV) and 9 (SEQ) are removed; ranks only remain.
taxa <- as.matrix(taxa)
taxa <- taxa[, -c(8:9)]

# Save taxonomy table (rows = ASVs, columns = Kingdom through Species)
write.table(taxa, "Tables/taxonomy_table_PROJECT_NAME.txt",
            sep = "\t", quote = FALSE, row.names = TRUE, col.names = TRUE)

# --- Step 4: build the abundance (OTU) table ---
# seqtab.nochim has rows = samples, columns = sequences.
# Transpose so rows = sequences (ASVs), then rename by position using correspondence.
# Position-based renaming is safe: assignTaxonomy always returns rows in the same order as the input seqtab.nochim columns.
otu <- t(seqtab.nochim)                       # rows = sequences, cols = samples
rownames(otu) <- correspondence$ASV            # rename by position — safe, no string matching needed
otu <- as.matrix(otu)
class(otu) <- "numeric"

# Save abundance table (rows = ASVs, columns = samples)
write.table(otu, "Tables/abundance_table_PROJECT_NAME.txt",
            sep = "\t", quote = FALSE, row.names = TRUE, col.names = TRUE)

cat("Taxonomy table:    Tables/taxonomy_table_PROJECT_NAME.txt\n")
cat("Abundance table:   Tables/abundance_table_PROJECT_NAME.txt\n")
cat("Correspondence:    Tables/correspondence_table_PROJECT_NAME.txt\n")
cat("ASVs exported:", nrow(otu), "\n")
cat("Samples:      ", ncol(otu), "\n")
```
